Integration of Multiple Constraints in ACG

### 2013-09-22 lenls/paper.tex
LENLS: Final proofreading

A presupposed condition to actually using an ACG to describe discourse in a
large scope is to have a large scale ACG grammar in the first place. The work
presented here was motivated by this quest for a wide-coverage ACG grammar, as
seen from a point of view of language as a system of largely orthogonal
constraints. Encoding ancillary constraints in a type system can make the
system hard to read and understand. If multiple constraints written in the
style of \cite{pogodalla2012controlling} are to be enforced in the same
grammar, we advocate for extending the formalism to prevent the incidental
complexity that would otherwise appear.

We begin with a short review of Abstract Categorial Grammars where we
highlight the ways in which Abstract Categorial Grammars facilitate
decomposition of linguistic description. We follow that by a closer look at an
example of a linguistic constraint. This discussion motivates our proposed
extension. We then present the Graphical Abstract Categorial Grammars and
consider some of its formal properties. We then finish with an illustration of
our approach which incorporates linguistic contribution from multiple sources
into a single coherent and simple grammar and examine the limitations and
challenges of adopting this approach.

@@

Now we have enough machinery in hand to describe how ACGs define
languages. Given some signature and a distinguished type, we can talk of the
\emph{abstract language} which is the set of well-typed terms built on that
signature having that distinguished type. If we then also consider a lexicon
and its abstract signature, then the language we get by mapping the abstract
language of the signature through that lexicon is called an \emph{object
  language}.

@@

The first constitutes the Constraint pattern used throughout
\cite{pogodalla2012controlling}. Imagine we want our grammar to enforce a
particular linguistic constraint, say that tensed clauses form islands for
quantifier scope, and suppose we want to proceed by refining our types so as
to make linguistic structures violating the constraint untypable. We could
dive into our grammar and rewrite it to use the new types. However, we can
also keep the existing grammar intact and complement it with a signature and a
lexicon that translates that signature to the abstract-most signature of our
existing grammar. Usually, the new signature will resemble the prior
abstract-most one, albeit with more fine-grained types necessary to express
the constraint in question.

@@

This pattern was used in \cite{pogodalla2007generalizing} to elaborate the
translation from the common Montague-style \cite{montague1973proper}
higher-order signature to its yield by introducing lower-order syntactic
structures more reminiscent of TAG or CFG trees. This helps make a clear
distinction between ambiguities caused by syntax and by semantics by
segregating the purely syntactic description in the $Syntax$ signature from
the issues of operator scope that have to be solved in the $SyntSem$ signature
(Figure~\ref{fig:acg-comp-sem}). However, in classical ACGs, we cannot exploit
this pattern as much as we would like. If we were to use this grammar as a
kernel on top of which we would like to express a syntactic constraint, we
would not be able to attach the constraint's signature directly to the
$Syntax$ node but we would instead need to attach it to
$SyntSem$\footnote{This is due to the fact that ACG diagrams are always
  arborescences, in which the root node represents the abstract language from
  whose terms are generated terms of all the object languages.}. This means
that the syntactic constraint would end up being expressed by strengthening
the type system of $SyntSem$, which only adds unnecessary complexity.

@@

In the above, we elaborate the types $NP$ and $N$ with features ($_{NEG{=}F}$
and $_{NEG{=}T}$) and we give the different types for the preposition \emph{de}
in the fragment for negation ($N$). The types accept an $NP$ and an $N$ as
arguments with any combination of values for the feature $NEG$, except for the
case when both the prepositional $NP$ and the $N$ being modified both contain
free negative determiners (i.e. there is no $N_{de_4} : NP_{NEG{=}T} \limp
N_{NEG{=}T} \limp N\_NEG{=}\ldots$). This encodes a constraint that there can
be only one free negative determiner per phrase (free as in not hidden inside
an embedded clause). Besides this constraint, the types also dictate how the
information about negative determiners should propagate from the argument
constituent to the complex constituent (in this case, it is simple
disjunction).

@@

We define a \emph{graphical abstract categorial grammar} as a directed acyclic
graph (DAG) whose nodes are labeled with signatures (and distinguished types)
and whose edges are labeled with lexicons, in other words a mathematical
reification of an ACG diagram that has been generalized from
arborescences to DAGs. We then search for an appropriate semantics for these
structures, i.e. how to determine what languages are defined by a graphical
ACG.

@@

G-ACG extrinsic languages correspond to ACG object languages with
intersection. More formally, while ACG object languages are ACG abstract
languages closed on transformation by a lexicon, G-ACG extrinsic languages are
ACG abstract languages closed on transformation by a lexicon and
intersection\footnote{Can be show by induction on the topological ordering of
  any given graphical ACG.}.

@@

We can observe another peculiarity on a more practical example. Consider the
G-ACG in Figure~\ref{fig:gacg}. In classical ACGs, one can always take an
element in a signature, like $Sem$, and by transduction find its corresponding
element in an another signature, like $Syntax$. However, in an G-ACG such as
this it is possible that the $Syntax$ term we obtain by transduction does not
belong to $\mathcal{E}_{\mathcal{G}}(Syntax)$ because it lacks antecedents in
either $Constr_1$ or $Constr_2$. This means that the notion of an extrinsic
language is not capable to give us the set of all meaning terms in $Sem$ which
actually correspond to syntactically correct sentences in this G-ACG.

@@

If we compare this new interpretation of G-ACGs to the former one, we find out
that in the case when the graph of the grammar is an arborescence they are
actually equivalent. This means that in classical ACGs, where all the diagrams
are arborescences, the two metaphors (nodes as languages and nodes as terms)
can be and are used interchangeably. However, as we start to consider
non-arborescent graphs, we find, interestingly, that the two paradigms diverge
(i.e. $\exists \mathcal{G}, u.\ \mathcal{E}_{\mathcal{G}}(u) \neq
\mathcal{P}_{\mathcal{G}}(u)$).

@@

Pangraphical languages turn out to be at least as expressive as the extrinsic
languages\footnote{This might come as no surprise given the extended domain of
  constraints compared to extrinsic languages (constrained by the entire graph
  as opposed to just the node's predecessors).}. The proof is carried out by
transforming a G-ACG such that a particular node will have the same
pangraphical language in the latter G-ACG as the extrinsic language it had in
the former. The nodes of the newly constructed G-ACG correspond to paths in
the former one terminating in the node in question\footnote{This means that
  when the G-ACG is $\mathcal{D}$, the diamond grammar, and the node in
  question is $Bottom$, then we will have two different nodes for the two
  paths $[Top, Left, Bottom]$ and $[Top, Right, Bottom]$.}. Thus the labelling
of nodes with terms which certifies a given term's presence in the
pangraphical language of the new G-ACG also serves as a proof of its presence
in the extrinsic language of the old G-ACG.

@@

We start with the extraction constraints. \cite{pogodalla2012controlling}
describes three different constraints on extraction, all of them expressed
using the Constraint pattern (Figure~\ref{fig:single-cont}). We can take the
union of these three G-ACGs to get the G-ACG on Figure~\ref{fig:merged-conts}.
Next, we can incorporate knowledge from \cite{pogodalla2007generalizing}, by
splitting the $\mathcal{L}_{Syn}$ lexicon and introducing a new intermediate
signature (Figure~\ref{fig:conts-elab}). The new signature, $Syntax$, deals
purely with syntax without any issues of operator scope and its functions have
lower-order types than those in $Syn$. We will finally add our contribution,
the signatures $Neg$ and $Agr$ and the lexicons $\mathcal{L}_{Neg}$ and
$\mathcal{L}_{Agr}$ which implement the constraint on negation we discussed
before and some simple notion of agreement, respectively.

@@

We have decomposed our grammar into 9 signatures and 8 lexicons. Besides
making the grammar more readable and easily navigable, it also makes the
grammar more attractive from a computational point of view. The complexity of
parsing the highest abstract-most structures and verifying all the constraints
is decomposed into smaller subtasks. A string can be parsed into its
representation at the $Syntax$ level fairly easily (compared to parsing it at
the higher levels) since $Syntax$ employs types of lower orders than the more
abstract signatures. Finding the syntactic structure for a string is thus
feasible. We then proceed to verifying the syntactic constraints. Here we can
observe that the lexicons $\mathcal{L}_{Neg}$ and $\mathcal{L}_{Agr}$ are mere
relabelings and therefore parsing (i.e. constraint verification) is trivial
(i.e. decidable). Now in order to determine the meaning of our sentence, we
have to continue parsing by finding an antecedent in $Syn$. This involves
higher-order types used to encode the various scope-bearing
operators. Finally, we can opt in to verify the high-level syntax-semantics
constraints which employ dependent types. However, the lexicons in this case
resemble relabelings and a semi-decidable procedure could be used to solve the
problem. Having the grammar decomposed like this opens it up for computational
processing by letting us engage in different tactics at different stages
instead of having to parse a single hidden structure which enforces
grammaticality w.r.t. every constraint. G-ACGs share these parks also with
classical ACGs but they let use this composition in another dimension, which
facilitates the writing of multiple constraints.

@@

Interesting avenues of future work include: digging deeper into the
language-theoretical properties of G-ACGs (Are ACG object languages closed on
intersection? Are extrinsic languages as expressive as pangraphical
languages?), searching for a metagrammatical description of constraint
signatures to ease work on G-ACG grammars and finally building large-scale
grammars to verify the usability of our approach.

>>>>

A presupposed condition to actually using an ACG to describe discourse in a
large scope is to have a large scale ACG grammar in the first place. The work
presented here was motivated by this quest for a wide-coverage ACG grammar, as
seen from a point of view of language as a system of largely orthogonal
constraints. Encoding ancillary constraints in a type system can make the
system hard to read and understand. If multiple constraints written in the
style of \cite{pogodalla2012controlling} are to be enforced in the same
grammar, we advocate for extending the formalism to prevent the incidental
complexity that would otherwise emerge.

We begin with a short review of Abstract Categorial Grammars where we
highlight the ways in which Abstract Categorial Grammars facilitate
decomposition of linguistic description. We follow that by a closer look at an
example of a linguistic constraint. This discussion motivates our proposed
extension. We then present the Graphical Abstract Categorial Grammars and
consider some of its formal properties. We then finish with an illustration of
our approach which incorporates linguistic contributions from multiple sources
into a single coherent and simple grammar and examine the limitations and
challenges of adopting this approach.

@@

Now we have enough machinery in hand to describe how ACGs define
languages. Given some signature and a distinguished type, we can talk of the
\emph{abstract language}, which is the set of well-typed terms built on the
signature and having the distinguished type. If we then also consider a
lexicon and its abstract signature, then the language we get by mapping the
abstract language of the signature through the lexicon is called an
\emph{object language}.

@@

The first constitutes the Constraint pattern used throughout
\cite{pogodalla2012controlling}. Imagine we want our grammar to enforce a
particular linguistic constraint, say that tensed clauses form islands for
quantifier scope, and suppose we want to proceed by refining our types so as
to make linguistic structures violating the constraint untypable. We could
dive into our grammar and rewrite it to use the new types. However, we can
also keep the existing grammar intact and complement it with a signature and a
lexicon that translates that signature to the abstract-most signature of our
existing grammar (Figure~\ref{fig:acg-comp-constr}). Usually, the new
signature will resemble the prior abstract-most one, albeit with more
fine-grained types necessary to express the constraint in question.

@@

This pattern was used in \cite{pogodalla2007generalizing} to elaborate the
translation from the common Montague-style \cite{montague1973proper}
higher-order signature to its yield by introducing lower-order syntactic
structures more reminiscent of TAG or CFG trees. This helps make a clear
distinction between ambiguities caused by syntax and by semantics by
segregating the purely syntactic description in the $Syntax$ signature from
the issues of operator scope that have to be solved in the $SyntSem$ signature
(Figure~\ref{fig:acg-comp-sem}). However, in classical ACGs, we cannot exploit
this pattern as much as we would like. If we were to use this grammar as a
kernel on top of which we would like to express a syntactic constraint, we
would not be able to attach the constraint's signature directly to the
$Syntax$ node but we would instead need to attach it to
$SyntSem$\footnote{This is due to the fact that ACG diagrams are always
  arborescences, in which the root node represents the abstract language from
  whose terms are generated the terms of all the object languages.}. This
means that the syntactic constraint would end up being expressed in terms of
the types of the syntax-semantics interface ($SyntSem$) rather than just pure
syntax ($Syntax$), which only adds unnecessary complexity.

@@

In the above, we elaborate the types $NP$ and $N$ with features ($NEG{=}F$
and $NEG{=}T$) and we give the different types for the preposition \emph{de}
in the fragment for negation ($N$). The types accept an $NP$ and an $N$ as
arguments with any combination of values for the feature $NEG$, except for the
case when both the prepositional $NP$ and the $N$ being modified both contain
free negative determiners (i.e. there is no $N_{de_4} : NP_{NEG{=}T} \limp
N_{NEG{=}T} \limp N_{NEG{=}}\ldots$). This encodes a constraint that there can
be only one free negative determiner per phrase (free as in not hidden inside
an embedded clause). Besides this constraint, the types also dictate how the
information about negative determiners should propagate from the argument
constituent to the complex constituent (in this case, it is simple
disjunction).

@@

We define a \emph{graphical abstract categorial grammar} as a directed acyclic
graph (DAG) whose nodes are labeled with signatures (and distinguished types)
and whose edges are labeled with lexicons, in other words, a mathematical
reification of an ACG diagram that has been generalized from arborescences to
DAGs. We then search for an appropriate semantics for these structures,
i.e. how to determine what languages are defined by a graphical ACG.

@@

G-ACG extrinsic languages correspond to ACG object languages with
intersection. More formally, while ACG object languages are ACG abstract
languages closed on transformation by a lexicon, G-ACG extrinsic languages are
ACG abstract languages closed on transformation by a lexicon and
intersection\footnote{Can be shown by induction on the topological ordering of
  any given graphical ACG.}.

@@

We can observe another peculiarity on a more practical example. Consider the
G-ACG in Figure~\ref{fig:gacg}. In classical ACGs, one can always take an
element in a signature, like $Sem$, and by transduction find its corresponding
element in an another signature, like $Syntax$. However, in a G-ACG such as
this it, is possible that the $Syntax$ term we obtain by transduction does not
belong to $\mathcal{E}_{\mathcal{G}}(Syntax)$ because it lacks antecedents in
either $Constr_1$ or $Constr_2$. This means that the notion of an extrinsic
language is not capable to give us the set of all meaning terms in $Sem$ which
actually correspond to syntactically correct sentences in this G-ACG.

@@

If we compare this new interpretation of G-ACGs to the former one, we find out
that in the case when the graph of the grammar is an arborescence, they are
actually equivalent. This means that in classical ACGs, where all the diagrams
are arborescences, the two metaphors (nodes as languages and nodes as terms)
can be, and are, used interchangeably. However, as we start to consider
non-arborescent graphs, we find, interestingly, that the two paradigms diverge
(i.e. $\exists \mathcal{G}, u.\ \mathcal{E}_{\mathcal{G}}(u) \neq
\mathcal{P}_{\mathcal{G}}(u)$).

@@

Pangraphical languages turn out to be at least as expressive as the extrinsic
languages\footnote{This might come as no surprise given the extended domain of
  constraints compared to extrinsic languages (constrained by the entire graph
  as opposed to just the node's predecessors).}. The proof is carried out by
transforming a G-ACG such that a particular node will have the same
pangraphical language in the latter G-ACG as the extrinsic language it had in
the former. The nodes of the newly constructed G-ACG correspond to paths in
the former one terminating in the node in question\footnote{This means that
  when the G-ACG is $\mathcal{D}$, the diamond grammar, and the node in
  question is $Bottom$, then we will have two different nodes for the two
  paths $[Top, Left, Bottom]$ and $[Top, Right, Bottom]$.}. Thus the labelling
of nodes with terms which certifies a given term's presence in the
pangraphical language of the new G-ACG also serves as a proof of its presence
in the extrinsic language of the old G-ACG and vice versa.

@@

We start with the extraction constraints. \cite{pogodalla2012controlling}
describes three different constraints on extraction, all of them expressed
using the Constraint pattern (Figure~\ref{fig:single-cont}). We can take the
union of these three G-ACGs to get the G-ACG on Figure~\ref{fig:merged-conts}.
Next, we can incorporate knowledge from \cite{pogodalla2007generalizing}, by
splitting the $\mathcal{L}_{Syn}$ lexicon and introducing a new intermediate
signature (Figure~\ref{fig:conts-elab}). The new signature, $Syntax$, deals
purely with syntax without any issues of operator scope and its functions have
lower-order types than those in $Syn$. We will finally add our contribution,
the signatures $Neg$ and $Agr$ and the lexicons $\mathcal{L}_{Neg}$ and
$\mathcal{L}_{Agr}$ which implement the constraint on negation we discussed
before and some simple notion of agreement, respectively
(Figure~\ref{fig:final-synth}).

@@

We have decomposed our grammar into 9 signatures and 8 lexicons. Besides
making the grammar more readable and easily navigable, it also makes the
grammar more attractive from a computational point of view. The complexity of
parsing the highest abstract-most structures and verifying all the constraints
is decomposed into smaller subtasks. A string can be parsed into its
representation at the $Syntax$ level fairly easily (compared to parsing it at
the higher levels) since $Syntax$ employs types of lower orders than the more
abstract signatures. Finding the syntactic structure for a string is thus
feasible. We then proceed to verifying the syntactic constraints. Here we can
observe that the lexicons $\mathcal{L}_{Neg}$ and $\mathcal{L}_{Agr}$ are mere
relabelings and therefore parsing (i.e. constraint verification) is trivial
(i.e. decidable). Now in order to determine the meaning of our sentence, we
have to continue parsing by finding an antecedent in $Syn$. This involves
higher-order types used to encode the various scope-bearing
operators. Finally, we can opt in to verify the high-level syntax-semantics
constraints which employ dependent types. However, the lexicons in this case
resemble relabelings and a semi-decidable procedure could be used to solve the
problem. Having the grammar decomposed like this opens it up for computational
processing by letting us engage in different tactics at different stages
instead of having to parse a single hidden structure which enforces
grammaticality w.r.t. every constraint. G-ACGs share these perks also with
classical ACGs but they let use this composition in another dimension, which
facilitates the writing of multiple constraints.

@@

Interesting avenues of future work include: digging deeper into the
language-theoretical properties of G-ACGs (Are ACG object languages closed on
intersection? Are extrinsic languages as expressive as pangraphical
languages?), searching for a metagrammatical description of constraint
signatures to ease work on G-ACG grammars, and finally building large-scale
grammars to verify the usability of our approach.
